### 2022-03-25 paper/localpsf.tex
fix typo

Let $\febasis_1, \febasis_2, \dots, \febasis_\fedim$ be a set of continuous Galerkin finite element basis functions used to discretize the problem on a mesh with mesh size parameter $h$, let $V_h := \Span\left(\febasis_1, \febasis_2, \dots, \febasis_\fedim\right)$ be the corresponding finite element space under the $L^2$ inner product, and let $p_i \in \mathbb{R}^\gdim$, $i=1,\dots, \fedim$ be the Lagrange nodes associated with the functions $\febasis_i$. We assume that the finite element basis satisfies the Kronecker property, i.e., $\febasis_i(p_i)=1$ and $\febasis_i(p_j)=0$ if $i\neq j$. 
For $u_h \in V_h \subset L^2(\Omega)$ we write $\bm{u} \in \mathbb{R}^N_\mathbf{M}$ to denote the coefficient vector for $u_h$ with respect to the finite element basis $\febasis_1, \dots, \febasis_\fedim$, i.e.,
\begin{equation}
\label{eq:fem_coeff_basis}
u_h(x) = \sum_{i=1}^\fedim \bm{u}_i \febasis_i(x).
\end{equation}
Linear functionals $\rho_h \in V_h' \supset L^2(\Omega)'$ have coefficient dual vectors $\boldsymbol{\rho}\in \mathbb{R}^N_{\mathbf{M}^{-1}}$, with entries 
\begin{equation*}
\left(\boldsymbol{\rho}_h\right)_i = \rho_h(\febasis_i), \qquad i=1,\dots,N. 
\end{equation*}
Here $\bm{M} \in \mathbb{R}^{\fedim \times \fedim}$ denotes the sparse finite element mass matrix which has entries
\begin{equation*}
\bm{M}_{ij}=\int_\Omega \febasis_i(x) \febasis_j(x) dx, \qquad i,j=1,\dots,N,
\end{equation*}
the space $\mathbb{R}^N_\mathbf{M}$ is $\mathbb{R}^N$ with the inner product $(\mathbf{u},\mathbf{v})_\mathbf{M} := \mathbf{u}^T \mathbf{M} \mathbf{v}$, and $\mathbb{R}^N_{\mathbf{M}^{-1}}$ is the analogous space with $\mathbf{M}^{-1}$ replacing $\mathbf{M}$. Direct calculation shows that $\mathbf{R}^N_\mathbf{M}$ and $\mathbf{R}^N_{\mathbf{M}^{-1}}$ are isomorphic to $V_h$ and $V_h'$ as Hilbert spaces, respectively.
%, in the sense that $\left(u_h, v_h\right)_{L^2(\Omega)} = \left(\mathbf{u}, \mathbf{v}\right)_\mathbf{M}$, $\left(\rho_h, \nu_h\right)_{L^2(\Omega)} = \left(\boldsymbol{\rho}, \boldsymbol{\nu}\right)_{\mathbf{M}^{-1}}$, and $\rho_h(u) = \boldsymbol{\rho}^T\mathbf{u}$, for $u_h,v_h \in V_h$ and $\rho_h, \nu_h \in V_h'$.

After discretization, the operator $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ is replaced by an operator $A_h:V_h \rightarrow V_h'$, which becomes an operator
\begin{equation*}
\mathbf{A}:\mathbb{R}^N_\mathbf{M} \rightarrow \mathbb{R}^N_{\mathbf{M}^{-1}}
\end{equation*}
under the isomorphism discussed above. Our method is agnostic to the computational procedure for approximating $\Aop$ with $\mathbf{A}$. What is important is that we do not have direct access to matrix entries $\mathbf{A}_{ij}$. Rather, we have a computational procedure that allows us to compute matrix-vector products $\mathbf{u}\mapsto \mathbf{A}\mathbf{u}$, and computing these matrix-vector products is costly.
%, and each of these matrix-vector products is costly to compute. 
%For example, some computational procedures possess the Galerkin property, in which case $\mathbf{A}_{ij} = \left(\Aop \febasis_j\right)(\febasis_i)$. 
%Our approximation method is agnostic to the computational procedure for performing these matrix-vector products. 
%What is important is that $\mathbf{A}$ is dense, only available via matrix-vector products, and each of these matrix-vector products is costly to compute. 
Of course, matrix entries can be computed via a matrix-vector products as $\mathbf{A}_{ij} = \left(\mathbf{A}\mathbf{e}_j\right)_i$,
%\begin{equation*}
%\mathbf{A}_{ij} = \left(\mathbf{A}\mathbf{e}_j\right)_i,
%\end{equation*}
where $\mathbf{e}_j=(0,\dots,0,1,0,\dots,0)^T$ is the length $N$ unit vector with one in the $j$th coordinate and zeros elsewhere. But computing the matrix-vector product $\mathbf{e}_j \mapsto \mathbf{A}\mathbf{e}_j$ is costly, and therefore wasteful if we do not use other matrix entries in the $j$th column of $\mathbf{A}$. Hence, methods for approximating $\mathbf{A}$ are computationally intractable if they require accessing scattered matrix entries from many different rows and columns of $\mathbf{A}$. 
The operator $A_h:V_h \rightarrow V_h'$ can be written in integral kernel form as
\begin{equation}
(A_h u_h)(v_h) := \int_\Omega \int_\Omega v_h(y) \Phi_h(y,x) u_h(x) dx dy
\end{equation}
for some kernel $\Phi_h$ which may differ from $\Phi$ due to discretization error. Since the functions $V_h$ are continuous at $x$, the delta distribution $\delta_x$ is a continuous linear functional on $V_h$, which has a discrete dual vector $\boldsymbol{\delta}_x \in \mathbb{R}^N_{\mathbf{M}^{-1}}$ with entries $\left(\boldsymbol{\delta}_x\right)_i = \febasis_i(x)$ for $i=1,\dots,N$. Additionally, it is straightforward to show that the Riesz representation, $\rho_h^* \in V_h$, of a functional $\rho \in V_h'$ has coefficient vector
\begin{equation*}
\boldsymbol{\rho}^* = \mathbf{M}^{-1} \boldsymbol{\rho}.
\end{equation*}
Therefore, the formula for the impulse response from \eqref{eq:impulse_response_delta_action} becomes 
\begin{equation}
\label{eq:discrete_kernel}
\boldsymbol{\impulseresponse}_x = \left(A_h \delta_x^*\right)^* =  \mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1} \boldsymbol{\delta}_x,
\end{equation}
and the $(y,x)$ kernel entry of $\Phi_h$ may be written as
\begin{equation}
\label{eq:approx_kernel_entry_81}
	\Phi_h(y,x) = \boldsymbol{\delta}_y^T \boldsymbol{\impulseresponse}_x = \boldsymbol{\delta}_y^T \mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1} \boldsymbol{\delta}_x.
\end{equation}
Let $\mathbf{\Phi} \in \mathbb{R}^{\fedim \times \fedim}$ be the following dense matrix of kernel entries evaluated at all pairs of Lagrange nodes:
\begin{equation}
\label{eq:Akerpcmat_entries}
\mathbf{\Phi}_{ij} := \Phi_h(p_i, p_j).
\end{equation}
Because of the Kronecker property of the finite element basis, we have $\boldsymbol{\delta}_{p_i} = \mathbf{e}_i$. Thus from \eqref{eq:approx_kernel_entry_81} we have
$
	\Phi_h(p_i,p_j) = \left(\mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1}\right)_{ij},
$
which implies
\begin{equation}
\label{eq:boldA}
	\mathbf{A} = \bm{M} \mathbf{\Phi} \bm{M}.
\end{equation}
Broadly, our method will construct an H-matrix approximation of $\mathbf{A}$ by forming H-matrix approximations of $\boldsymbol{\Phi}$ and $\mathbf{M}$ (or a lumped mass version of $\mathbf{M}$), then multiplying these matrices per \eqref{eq:boldA}. The central challenge is that classical H-matrix construction methods require access to arbitrary matrix entries $\mathbf{\Phi}_{ij}$, but these matrix entries are not easily accessible. The bulk of our method is therefore dedicated to forming approximations of these matrix entries that can be evaluated rapidly.

@@

The impulse response $\impulseresponse_x$ may be interpreted as a scaled probability distribution because of the non-negative integral kernel property. Let $\spatialvol:\Omega \rightarrow \mathbb{R}$,
\begin{equation*}
\spatialvol(x) := \int_\Omega \impulseresponse_x(y) dy
\end{equation*}
denote the spatially varying scaling factor, and let $\mu:\Omega \rightarrow \mathbb{R}^d$, $\Sigma:\Omega \rightarrow \mathbb{R}^{d \times d},$
\begin{equation*}
\spatialmean(x) := \mathbb{E}(\impulseresponse_x / V(x)), \qquad
\spatialcov(x) := \Var(\impulseresponse_x /V(x)),
\end{equation*}
to be the spatially varying mean and covariance of the normalized version of $\impulseresponse_x$, respectively. Imagine $\phi_x$ as a mound of dirt piled on a localized region in $\Omega$, with $\phi_x(y)$ being the height of the dirt mound at location $y$. The scalar $V(x)$ is the total volume of dirt in the mound, the vector $\mu(x)$ is the center of mass of the dirt mound, and the matrix $\Sigma(x)$ contains information about how wide the dirt mound is in each spatial direction.

>>>>

Let $\febasis_1, \febasis_2, \dots, \febasis_\fedim$ be a set of continuous Galerkin finite element basis functions used to discretize the problem on a mesh with mesh size parameter $h$, let $V_h := \Span\left(\febasis_1, \febasis_2, \dots, \febasis_\fedim\right)$ be the corresponding finite element space under the $L^2$ inner product, and let $p_i \in \mathbb{R}^\gdim$, $i=1,\dots, \fedim$ be the Lagrange nodes associated with the functions $\febasis_i$. We assume that the finite element basis satisfies the Kronecker property, i.e., $\febasis_i(p_i)=1$ and $\febasis_i(p_j)=0$ if $i\neq j$. 
For $u_h \in V_h \subset L^2(\Omega)$ we write $\bm{u} \in \mathbb{R}^N_\mathbf{M}$ to denote the coefficient vector for $u_h$ with respect to the finite element basis $\febasis_1, \dots, \febasis_\fedim$, i.e.,
\begin{equation}
\label{eq:fem_coeff_basis}
u_h(x) = \sum_{i=1}^\fedim \bm{u}_i \febasis_i(x).
\end{equation}
Linear functionals $\rho_h \in V_h' \supset L^2(\Omega)'$ have coefficient dual vectors $\boldsymbol{\rho}\in \mathbb{R}^N_{\mathbf{M}^{-1}}$, with entries 
\begin{equation*}
\boldsymbol{\rho}_i = \rho_h(\febasis_i), \qquad i=1,\dots,N. 
\end{equation*}
Here $\bm{M} \in \mathbb{R}^{\fedim \times \fedim}$ denotes the sparse finite element mass matrix which has entries
\begin{equation*}
\bm{M}_{ij}=\int_\Omega \febasis_i(x) \febasis_j(x) dx, \qquad i,j=1,\dots,N,
\end{equation*}
the space $\mathbb{R}^N_\mathbf{M}$ is $\mathbb{R}^N$ with the inner product $(\mathbf{u},\mathbf{v})_\mathbf{M} := \mathbf{u}^T \mathbf{M} \mathbf{v}$, and $\mathbb{R}^N_{\mathbf{M}^{-1}}$ is the analogous space with $\mathbf{M}^{-1}$ replacing $\mathbf{M}$. Direct calculation shows that $\mathbf{R}^N_\mathbf{M}$ and $\mathbf{R}^N_{\mathbf{M}^{-1}}$ are isomorphic to $V_h$ and $V_h'$ as Hilbert spaces, respectively.
%, in the sense that $\left(u_h, v_h\right)_{L^2(\Omega)} = \left(\mathbf{u}, \mathbf{v}\right)_\mathbf{M}$, $\left(\rho_h, \nu_h\right)_{L^2(\Omega)} = \left(\boldsymbol{\rho}, \boldsymbol{\nu}\right)_{\mathbf{M}^{-1}}$, and $\rho_h(u) = \boldsymbol{\rho}^T\mathbf{u}$, for $u_h,v_h \in V_h$ and $\rho_h, \nu_h \in V_h'$.

After discretization, the operator $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ is replaced by an operator $A_h:V_h \rightarrow V_h'$, which becomes an operator
\begin{equation*}
\mathbf{A}:\mathbb{R}^N_\mathbf{M} \rightarrow \mathbb{R}^N_{\mathbf{M}^{-1}}
\end{equation*}
under the isomorphism discussed above. Our method is agnostic to the computational procedure for approximating $\Aop$ with $\mathbf{A}$. What is important is that we do not have direct access to matrix entries $\mathbf{A}_{ij}$. Rather, we have a computational procedure that allows us to compute matrix-vector products $\mathbf{u}\mapsto \mathbf{A}\mathbf{u}$, and computing these matrix-vector products is costly.
%, and each of these matrix-vector products is costly to compute. 
%For example, some computational procedures possess the Galerkin property, in which case $\mathbf{A}_{ij} = \left(\Aop \febasis_j\right)(\febasis_i)$. 
%Our approximation method is agnostic to the computational procedure for performing these matrix-vector products. 
%What is important is that $\mathbf{A}$ is dense, only available via matrix-vector products, and each of these matrix-vector products is costly to compute. 
Of course, matrix entries can be computed via a matrix-vector products as $\mathbf{A}_{ij} = \left(\mathbf{A}\mathbf{e}_j\right)_i$,
%\begin{equation*}
%\mathbf{A}_{ij} = \left(\mathbf{A}\mathbf{e}_j\right)_i,
%\end{equation*}
where $\mathbf{e}_j=(0,\dots,0,1,0,\dots,0)^T$ is the length $N$ unit vector with one in the $j$th coordinate and zeros elsewhere. But computing the matrix-vector product $\mathbf{e}_j \mapsto \mathbf{A}\mathbf{e}_j$ is costly, and therefore wasteful if we do not use other matrix entries in the $j$th column of $\mathbf{A}$. Hence, methods for approximating $\mathbf{A}$ are computationally intractable if they require accessing scattered matrix entries from many different rows and columns of $\mathbf{A}$.

The operator $A_h:V_h \rightarrow V_h'$ can be written in integral kernel form as
\begin{equation}
(A_h u_h)(v_h) := \int_\Omega \int_\Omega v_h(y) \Phi_h(y,x) u_h(x) dx dy
\end{equation}
for some kernel $\Phi_h$ which may differ from $\Phi$ due to discretization error. Since the functions $V_h$ are continuous at $x$, the delta distribution $\delta_x$ is a continuous linear functional on $V_h$, which has a discrete dual vector $\boldsymbol{\delta}_x \in \mathbb{R}^N_{\mathbf{M}^{-1}}$ with entries $\left(\boldsymbol{\delta}_x\right)_i = \febasis_i(x)$ for $i=1,\dots,N$. Additionally, it is straightforward to show that the Riesz representation, $\rho_h^* \in V_h$, of a functional $\rho \in V_h'$ has coefficient vector
\begin{equation*}
\boldsymbol{\rho}^* = \mathbf{M}^{-1} \boldsymbol{\rho}.
\end{equation*}
Therefore, the formula for the impulse response from \eqref{eq:impulse_response_delta_action} becomes 
\begin{equation}
\label{eq:discrete_kernel}
\boldsymbol{\impulseresponse}_x = \left(A_h \delta_x^*\right)^* =  \mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1} \boldsymbol{\delta}_x,
\end{equation}
and the $(y,x)$ kernel entry of $\Phi_h$ may be written as
\begin{equation}
\label{eq:approx_kernel_entry_81}
	\Phi_h(y,x) = \boldsymbol{\delta}_y^T \boldsymbol{\impulseresponse}_x = \boldsymbol{\delta}_y^T \mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1} \boldsymbol{\delta}_x.
\end{equation}
Let $\mathbf{\Phi} \in \mathbb{R}^{\fedim \times \fedim}$ be the following dense matrix of kernel entries evaluated at all pairs of Lagrange nodes:
\begin{equation}
\label{eq:Akerpcmat_entries}
\mathbf{\Phi}_{ij} := \Phi_h(p_i, p_j).
\end{equation}
Because of the Kronecker property of the finite element basis, we have $\boldsymbol{\delta}_{p_i} = \mathbf{e}_i$. Thus from \eqref{eq:approx_kernel_entry_81} we have
$
	\Phi_h(p_i,p_j) = \left(\mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1}\right)_{ij},
$
which implies
\begin{equation}
\label{eq:boldA}
	\mathbf{A} = \bm{M} \mathbf{\Phi} \bm{M}.
\end{equation}
Broadly, our method will construct an H-matrix approximation of $\mathbf{A}$ by forming H-matrix approximations of $\boldsymbol{\Phi}$ and $\mathbf{M}$ (or a lumped mass version of $\mathbf{M}$), then multiplying these matrices per \eqref{eq:boldA}. The central challenge is that classical H-matrix construction methods require access to arbitrary matrix entries $\mathbf{\Phi}_{ij}$, but these matrix entries are not easily accessible. The bulk of our method is therefore dedicated to forming approximations of these matrix entries that can be evaluated rapidly.

@@

The impulse response $\impulseresponse_x$ may be interpreted as a scaled probability distribution because of the non-negative integral kernel property. Let $\spatialvol:\Omega \rightarrow \mathbb{R}$,
\begin{equation*}
\spatialvol(x) := \int_\Omega \impulseresponse_x(y) dy
\end{equation*}
denote the spatially varying scaling factor, and let $\mu:\Omega \rightarrow \mathbb{R}^d$, $\Sigma:\Omega \rightarrow \mathbb{R}^{d \times d},$
\begin{equation*}
\spatialmean(x) := \mathbb{E}(\impulseresponse_x / V(x)), \qquad
\spatialcov(x) := \Var(\impulseresponse_x /V(x)),
\end{equation*}
be the spatially varying mean and covariance of the normalized version of $\impulseresponse_x$, respectively. Imagine $\phi_x$ as a mound of dirt piled on a localized region in $\Omega$, with $\phi_x(y)$ being the height of the dirt mound at location $y$. The scalar $V(x)$ is the total volume of dirt in the mound, the vector $\mu(x)$ is the center of mass of the dirt mound, and the matrix $\Sigma(x)$ contains information about how wide the dirt mound is in each spatial direction.
